---
title: "Final_project_data_cleanup"
output: html_document
date: "2025-04-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

library needed packages
```{r, warning=FALSE, message=FALSE}
library("dplyr")
library("tidyverse")
library("ggplot2")
library("scales")
library("ggpubr")
library(glmnet)
library(ggrepel)

```

read in dataset
```{r}
data <- read.csv("Preprocessed_data.csv")
```

check the survey year, and only keep the data no earlier than 2013
```{r}
min(data$Survey_Year) # 1999
max(data$Survey_Year) # 2017
data_filtered <- data %>%
  filter(Survey_Year >= 2013)
dim(data_filtered) # 9549 160
```
# define plot theme
```{r}
pub_theme <- theme_classic() +
  theme(plot.title = element_text(size = 8)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(
    legend.title = element_text(size = 8),       # Title font
    legend.text = element_text(size = 6),        # Labels font
    legend.key.size = unit(0.3, "cm"),           # Box size
    legend.spacing.y = unit(0.1, "cm")           # Vertical spacing
  )
```


we will be using the blood pressure -hypertension definition defined by the 2017 ACC/AHA BP guideline.
For our  study, we will only be using the population that does have hypertension.

We will be using the blood pressure controlled defined by the 2017 ACC/AHA BP guideline.
We will change the variable to have 1 = controlled, 0 = not controlled and save it to a new variable bp_control_binary

```{r}
data_filtered <- data
data_filtered <- data_filtered %>%
  filter(htn_accaha == "Yes")
data_filtered$bp_control_binary <- ifelse(data_filtered$bp_control_accaha == "Yes", 1, 0)

```

#### data cleaning and imputation
We will then remove variables that have missing values over 50% and show the removed values
we will then impute numerical values with their median and categorical varibles with their mode
```{r}
# Calculate proportion of missing values per column
missing_prop <- colMeans(is.na(data_filtered))

# Identify columns to remove (more than 50% missing)
removed_vars <- names(missing_prop[missing_prop > 0.5])
#"chol_total"    "chol_hdl"      "chol_trig"     "chol_ldl"      "chol_nonhdl"  
# "LBDHDL"        "LBDLDL"        "FriedewaldLDL"
# Print the removed variable names
print(removed_vars)

data_filtered <- data_filtered[, missing_prop <= 0.5] # 153 variables left

```

imputation
```{r}
# impute variables, fill numeric with mean
#Convert all "missing" strings (empty, "NA", etc.) to proper NA
data_filtered[data_filtered == ""] <- NA
data_filtered[data_filtered == "NA"] <- NA
save(data_filtered, file = "unimputed_data.RData")


# impute variables, fill categorical with mode
# Define a mode function
Mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}
# Impute numeric and categorical columns
data_filtered <- data_filtered %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  mutate(across(where(~ is.factor(.) || is.character(.)), ~ {
    val <- .
    val[is.na(val)] <- Mode(val)
    return(val)
  }))

# Identify columns that have only one unique value
constant_cols <- sapply(data_filtered, function(x) length(unique(x[!is.na(x)])) <= 1)

# List of columns with only one unique value
constant_vars <- names(constant_cols[constant_cols == TRUE])

# Remove these constant columns from the dataset
data_filtered <- data_filtered[, !names(data_filtered) %in% constant_vars]

save(data_filtered, file = "cleaned_up_raw_data.RData")
```

```{r}
load("cleaned_up_raw_data.RData")
```


```{r}
set.seed(123)
# remove variables that are less meaningful, or leaking variables, or we know to have definite effects such as bp measurements, survey year...
data_for_reg <- data_filtered %>%
  select(-svy_weight_mec, -svy_strata, -svy_id, -bp_control_accaha, -bp_sys_mean, -bp_dia_mean, -svy_strata, -Begin_Year,  -YEAR, -BPXSY1,-BPXSY2, -BPXSY3, -BPXDI1, -BPXDI2, -BPXDI3,-SDDSRVYR, -demo_race, -demo_age_cat, -demo_gender, -svy)
# separate the data into data before and after 2013: increase and decrease phase
data_for_reg_inc <- data_for_reg %>% 
  filter(Survey_Year < 2013) %>% 
  select(-Survey_Year)
data_for_reg_dec <- data_for_reg %>% 
  filter(Survey_Year >= 2013) %>% 
  select(-Survey_Year)
# Add binary outcome variable AFTER dropping it
data_for_reg_inc$bp_control_binary <- as.factor(data_for_reg_inc$bp_control_binary)
data_for_reg_dec$bp_control_binary <- as.factor(data_for_reg_dec$bp_control_binary)
# scale numeric variables
data_for_reg_inc <- data_for_reg_inc %>%
  mutate(across(where(is.numeric), ~ scale(.)[, 1]))
data_for_reg_dec <- data_for_reg_dec %>%
  mutate(across(where(is.numeric), ~ scale(.)[, 1]))

# Convert all "Yes"/"No" in a dataframe to 1/0
data_for_reg_inc_cleaned <- data_for_reg_inc
data_for_reg_inc_cleaned[] <- lapply(data_for_reg_inc_cleaned, function(x) {
  if (is.character(x) || is.factor(x)) {
    x <- as.character(x)
    if (all(x %in% c("Yes", "No", NA))) {
      return(ifelse(x == "Yes", 1, ifelse(x == "No", 0, NA)))
    }
  }
  return(x)
})
data_for_reg_dec_cleaned <- data_for_reg_dec
data_for_reg_dec_cleaned[] <- lapply(data_for_reg_dec_cleaned, function(x) {
  if (is.character(x) || is.factor(x)) {
    x <- as.character(x)
    if (all(x %in% c("Yes", "No", NA))) {
      return(ifelse(x == "Yes", 1, ifelse(x == "No", 0, NA)))
    }
  }
  return(x)
})
```

```{r}
save(data_for_reg_dec_cleaned, data_for_reg_inc_cleaned,data_for_reg, file = "cleaned_data.RData")
```

